Boundary terms in the action for Causal Sets

### 2014-09-18 paper_v6.tex
initial commit

\newcommand{\IGNORE}[1]{}
\newcommand{\be}{\begin{equation}}
\newcommand{\ee}{\end{equation}}
\newcommand{\bea}{\begin{aligned}}
\newcommand{\eea}{\end{aligned}}
\newcommand{\mb}[1]{\marginnote{\texttt{\small MB:\,#1}}}
\renewcommand{\ij}[1]{\marginnote{\texttt{\small IJ:\,#1}}}
\newcommand{\ibar}{{\overline{\emph\i}}}
\newcommand{\jbar}{{\overline{\emph\j}}}

\title{Boundary terms in the action for Causal Sets}
 \author[a]{M. Buck}
 \author[a,b]{\!, F. Dowker}
 \author[a]{and I. Jubb\,}
\affiliation[a]{Theoretical Physics Group, Blackett Laboratory, Imperial College, London, SW7 2AZ, U.K.}
\affiliation[b]{Institute for Quantum Computing, University of Waterloo, ON, N2L 2Y5, Canada}

\abstract{ 
We propose a family of formulae for the boundary term in the action of a causal set that is well-approximated by a continuum manifold with spacelike boundary. %The boundary term is proportional to the difference in the number of elements immediately to future and the number of elements immediately to the past of the surface. 
We show that in the continuum limit one recovers the Gibbons-Hawking-York boundary term in the mean.
}

@@

We have performed analogous simulations for the second formula in~\eqref{GH_boundary_to_causet}, involving $N^{(1)}_{max}$ and $N^{(0)}_{max}$. In this case, heuristic arguments of the type given above are hampered by the fact that the two random variables whose difference is taken are not independent. However, the results for the scaling of the fluctuations turn out to be identical to those obtained in the first case. 
\newpage

>>>>

\newcommand{\IGNORE}[1]{}
\newcommand{\be}{\begin{equation}}
\newcommand{\ee}{\end{equation}}
\newcommand{\bea}{\begin{aligned}}
\newcommand{\eea}{\end{aligned}}
\newcommand{\mb}[1]{\marginnote{\texttt{\small MB:\,#1}}}
\renewcommand{\ij}[1]{\marginnote{\texttt{\small IJ:\,#1}}}
\newcommand{\ibar}{{\overline{\emph\i}}}
\newcommand{\jbar}{{\overline{\emph\j}}}
\newcommand{\mnote}[1]{\marginpar{\raggedright \scriptsize   #1 } }

\title{Boundary terms in the action for Causal Sets}
 \author[a]{Michel Buck}
 \author[a,b]{\!, Fay Dowker}
 \author[a]{ Ian Jubb\,}
 \author[c]{and Sumati Surya}
\affiliation[a]{Theoretical Physics Group, Blackett Laboratory, Imperial College, London, SW7 2AZ, U.K.}
\affiliation[b]{Institute for Quantum Computing, University of Waterloo, ON, N2L 2Y5, Canada}
\affiliation[c]{Raman Research Institute, CV Raman Ave, Sadashivanagar, Bangalore 560080, India}
\abstract{ 
We propose a family of formulae for the boundary term in the action of a causal set that is well-approximated by a continuum manifold with spacelike boundary. %The boundary term is proportional to the difference in the number of elements immediately to future and the number of elements immediately to the past of the surface. 
We show that in the continuum limit one recovers the Gibbons-Hawking-York boundary term in the mean.
}

@@

We have performed analogous simulations for the second formula in~\eqref{GH_boundary_to_causet}, involving $N^{(1)}_{max}$ and $N^{(0)}_{max}$. In this case, heuristic arguments of the type given above are hampered by the fact that the two random variables whose difference is taken are not independent. However, the results for the scaling of the fluctuations turn out to be identical to those obtained in the first case.

\section{A Candidate Boundary Term for an Alexandrov Interval}

Until now our approach has been to reconstruct the  continuum GHY term  from the information in a causal set. 
The candidate described in the previous sections for a spatial boundary has a natural expression in a causal set  in  terms of its maximal and minimal elements which are the  analogs of a  space-like boundary.  A suitable analog  of  a time-like boundary  term has been harder to  construct; there is no simple causal set  characterisation of a time-like boundary. Indeed, it seems difficult to characterise a causal set in a  spacetime region which is bounded by time-like surfaces. On the other hand, since past and future sets are natural in a causal set, such regions do find a natural characterisation in a causal set. An important example of such a region is the  Alexandrov interval which is used extensively  in calculations of causal set kinematics.  In the continuum such a region  is bounded by null hypersurfaces and their spatial joints.

Conversely in the continuum,   while the GHY term  is well defined for both a space-like and time-like boundary,  including their co-dimension 2 joints \cite{harris},  it is an open question what the contribution from a null  boundary  is \cite{ghh}. Apart from some sketchy suggestions in the literature \cite{neiman} there is no rigourous construction of such a term; indeed the standard variational route for finding the boundary contribution to the action appears to fail for null boundaries.

Given the lack of a  null GHY term, we can ask if causal set theory  might  in fact guide us in the right direction.   Our starting point is the $d$-dimensional  Benincasa-Dowker Action \cite{bd,gd} which is constructed from the mean number  $N_i^{(d)}$ \footnote{ $N_i^{(d)}$ is dimension  and geometry dependent, but for our purpose it suffices to retain the dimension label. } of $i+1$-element inclusive intervals in the causal set 
\begin{equation} 
8 \pi G {S_{BD}^d} = -\alpha_d \rho^{\frac{2-d}{d}}\biggl( N+ \frac{\beta_d}{ \alpha_d} \sum_{i=1}^{n_d-1}  C^{(d)}_{i} N _i^{(d)} \biggr) 
\label{bd} 
\end{equation}  
where   
\begin{equation} 
 \alpha_d= \left \{ %
\begin{array} {ccc}
-\frac{1}{\Gamma(\frac{2+d}{d})}c_d ^{2/d}  & \quad &  \mathrm{d} \, \, \mathrm{odd} \\
-2 \frac{1}{\Gamma(\frac{2+d}{d})}c_d ^{2/d} & \quad &  \mathrm{d \, \,  even} \\
\end{array} \right  
\quad \beta_d = \left \{ 
\begin{array} {ccc}
\frac{d+1}{2^{d-1}\Gamma(2/d +1)} c_d^{2/d}  & \quad &  \mathrm{d \, \, odd}\\ 
\frac{\Gamma(d/2+2)\Gamma(d/2+1)}{\Gamma(2/d)\Gamma(d)}c_d^{2/d}  & \quad & \mathrm{d \, \,  even}\\ 
\end{array} \right 
\end{equation} 
\begin{equation} 
n_d = \left \{ %
\begin{array} {ccc}
\frac{d-1}{2} + 2 & \quad &  \mathrm{d \, \, odd}\\ 
\frac{d}{2} + 2 & \quad & \mathrm{d \, \,  even}\\ 
\end{array} \right 
\end{equation} 
with $2^{-d/2} c_d$ is the volume of a unit Alexandrov interval $I(x,y)$, i.e., one for which the proper time $\tau(x,y)=1$. 
%%{ {\bf \color{red} this is not correct! $c_d= vol(S_{d-2}) \frac{ 2^{\frac{2-d}{2}}}{d(d-1)}$ and $vol(S_{d-2})$ the volume of the homogeneous $d-2$ sphere}}.  
The coefficients  of $N_i^{(d)}$ in the sum are are 
\begin{equation}
\label{cid}
 C_i^d= \left \{ %
\begin{array} {ccc}
\sum_{k=0}^{i-1} (-1)^k\binom{i-1}{k} \frac{\Gamma(d/2(k+1)+3/2)}{\Gamma(d/2+3/2) \Gamma(1+dk/2)} & \quad &  \mathrm{d \, \, odd}\\ 
\sum_{k=0}^{i-1} (-1)^k\binom{i-1}{k} \frac{\Gamma(d/2(k+1)+2)}{\Gamma(d/2+2) \Gamma(1+dk/2)} & \quad & \mathrm{d \, \,  even}\\ 
\end{array} \right 
\end{equation} 
These coefficients can be expressed more compactly as hypergeometric functions 
\begin{equation}
C_i^d={}_{q+1}F_{q}(\{ a_1, \ldots a_q, i-1\}, \{b_1, \ldots b_q \} |1)
\label{chyp} 
\end{equation} 
with $q=d/2+1/2, a_i=\frac{d+2i}{d}, b_i=2i/d$ for $d$ odd and $q=d/2, a_i=\frac{d+2i+2}{d}, b_i=2i/d$  for $d$ even.

If $S_{BD}^d$ were precisely the discrete analog of the Einstein Hilbert action {\it without} any boundary contributions, then in Minkowski spacetime  $S_{BD}^d$ should  vanish. However, as shown explicitly in \cite{bbdtwo} this is not the case for Alexandrov intervals in 2$d$  Minkowski spacetime. Indeed, there is a residual contribution that goes to a constant in the continuum limit $N\rightarrow \infty$.  When the action is evaluated in rectangular regions in the same spacetime or for Alexandrov intervals in  the flat trousers spacetime, this is no longer the case. Thus, it seems plausible that the 2$d$ BD action contains some boundary information. Moreover, in the large $N$ limit  the action goes to the same  non-vanishing constant irrespective of the interval size. While this might suggest a non-geometric or topological origin, we will now show that it is a part of a universal  behaviour for higher $d$ and in fact  has a geometrical origin.

Consider an Alexandrov interval $I(p,q)=I^+(p) \cap I^-(q)$ for two events $p,q$ in a causal spacetime.   While the  boundary of $I(p,q)$ may be quite complicated in general, some of these components must necessarily be null.  In Minkowski spacetime, for example,  the boundary of $I(p,q)$ consists of the two events $p$ and $q$, the subsets $B^\pm$ of the  two null hypersurfaces $\partial J^+(p)$ and $\partial J^-(q)$, and their intersection or ``joint''  $S_{d-2}=\partial J^+(p) \cap \partial J^-(q)$ which is a codimension 2 sphere.  Evaluating the BD action for an Alexandrov interval in flat spacetime should therefore give us terms that are only dependent on the boundary components.  In flat spacetime the $S_{d-2}$ joint  has radius $\tau(p,q)/2$, and thus increases with the size of the Alexandrov interval ${\mathrm{vol}}(I(p,q)) = N/\rho$, where $N$ is the mean of the number of causal set elements sprinkled into $I(p,q)$ . In what follows we take the continuum limit $\rho, N \rightarrow \infty$ while keeping ${\mathrm{vol}}(I(p,q))$ fixed. \mnote{\color{red} SS: It would be nice to have a fancy picture of an Alex. interval here..}  
%% If $p$ or $q$ were to be taken to time-like past or future infinity, this volume would go to infinity as expected.

In \cite{glasersurya} a closed form expression was obtained for the mean value of the number of $i+1$ element inclusive intervals  $N_i^d$ contained in an  Alexandrov interval in $d$ dimensional Minkowski  spacetime 
\begin{equation} 
N_i^d= \frac{\Gamma(d)^2}{\Gamma(i)} \sum_{k=0}^\infty \frac{(-1)^k}{\Gamma(k+1)} \frac{N^{k+i+2}}{(k+i+2)(k+i+1)} \frac{\Gamma(\frac{d(k+i)}{2}+1) \Gamma(\frac{d(k+i+1)}{2}+1)}{\Gamma(\frac{d(k+i)}{2} +d) \Gamma(\frac{d(k+i+1)}{2}+d)}, 
\end{equation}
where $i \geq 1$.   In taking the $N \rightarrow \infty$ limit, it   is convenient to consider the ratio
\begin{equation} 
\frac{ 8 \pi G S^d(C)}{vol(S^{d-2})} = \frac{b}{d(d-1) \Gamma(1+2/d) N^{\frac{d-2}{d}}} \left( N + \frac{\beta_d}{\alpha_d} \sum_{i=1}^{n_d} C_i^d N_i^d \right)  
\label{ratio} 
\end{equation} 
where $b=1$ for $d$ odd and $b=2$ for $d$ even. 
We use Mathematica to evaluate the $N \rightarrow \infty$  limit of this expression for $d=2, \ldots 9$. We aid the calculation by expanding the sum  $\sum_{i=1}^{n_d} C_i^d N_i^{(d)} = \sum_{j=1}^\infty A_j^d N^{j+1}$ as a  power series in  $N$. After a rearrangement and redefinition of indices we find that 
\begin{equation}
 A_j^d = \Gamma(d)^2\biggl( \sum_{i=1}^{D+2} (-1)^i \binom{j-1}{i-1}  C_i^d  \biggr) \frac{(-1)^j}{(j+1)!}\frac{\Gamma(\frac{d}{2}(j-1)+1)\Gamma(\frac{d}{2}j+1)}{\Gamma(\frac{d}{2}(j-1)+d) \Gamma(\frac{d}{2}j+d)}, 
\end{equation} 
where $d=2D$ for $d$ even and $d=2D+1$ for $d$ odd.   
% Using the abbreviated form $C_i^d=\sum_{l=0}^{i-1} \binom{i-1}{l}A(d,l)$ where $A(d,l)$ is defined using \ref{cid} 
% we find that 
% \begin{equation}
% \sum_{i=1}^{D+2} (-1)^i \binom{j-1}{i-1}  C_i^d = (-1)^{j-1} A(d,j-1),  
% \end{equation}      
% where we have used the binomial identities  BLAH. Thus 
% \begin{equation}
%  \sum_{i=1}^{n_d} C_i^d N_i^{(d)}= ((d-1)!)^2 \sum_{j=1}^\infty(-1)^j B(d,j) \frac{\Gamma(\frac{d}{2}j+1)}{\Gamma(j+2)\Gamma(\frac{d}{2}(j-1)+d)\Gamma(\frac{d}{2}j+d)}N^{j+1}
% \label{thesum} 
% \end{equation} 
% where 
% \begin{equation} 
% B(d,j) = \left \{ 
% \begin{array}{ccc} 
% \frac{\Gamma(\frac{d}{2}j+3/2)}{\Gamma(\frac{d}{2}+3/2)}  &\quad & d \, \, odd, \\  \frac{\Gamma(\frac{d}{2}j+2)}{\Gamma(\frac{d}{2}+2)}  &\quad & d \, \, even.  \\ 
% \end{array} 
% \right
% \end{equation} 
An explicit  calculation of $8\pi S_{BD}^d$ in $I(p,q)$ for $d=2, \ldots 9$  gives 
\begin{equation} 
\lim_{N \rightarrow \infty} \frac{ 8 \pi G S_{BD}^d(C)}{vol(S^{d-2})} = 1,  
\label{result} 
\end{equation}

\mnote{\color{red} SS: There are further simplifications I have that may help mathematica for $n>9$. It is still not clear}  Since $N_i^d$ can also be expressed as a hypergeometric function, using  (\ref{chyp}) it   is seems possible to evaluate (\ref{ratio}) using identities involving sums of products of such functions to get a closed form expression for arbitrary $d$.  In \cite{glasersurya} the asymptotic form of $N_i^d$ was explicitly calculated, and for $d>2$, the leading order was shown to  $~ N^{2-\frac{2}{d}}$ which dominates $N$ dependence for $d>2$. To be consistent with  (\ref{result}) this would mean that the leading order contributions from the $N_i^d$ cancel in the action. This is indeed explicitly seen to be the case, again for $d=3, \ldots 9$. Similarly, for $d=2$, where the leading order contribution is $~ N \log N$.  Indeed, in all cases the next to leading order terms given in \cite{glasersurya} also do not have the requisite $~N^{\frac{d-2}{d}}$ dependence required by (\ref{result}); in this case, the explicit coefficients are not available to   check that they give a vanishing contribution to (\ref{result}).  This suggests that  we cannot rely   on the the asymptotic form of the individual terms  in the sum  for an analytic proof for general $d$, which could therefore be quite non-trivial.

The result we have obtained is for flat spacetime. One may speculate that the presence of curvature will not modify the essential result, i.e., that the only boundary contribution is from the volume of the spatial $d-2$ sphere at the joint, at least in the case when the boundary consists of the same type of components.   What will change of course is the fact that the principal radii of the $d-2$ sphere could be different at  every point, hence complicating the relationship between the volume of the interval and the volume of this sphere. In  \cite{Khetrapal_Sumati:Causal_Diamond_Volume} it was shown that in an RNC, to first order in curvature, 
the intrinsic boundary geometry of the Alexandrov interval is still flat which would simplify this relationship. Using an RNC version of the calculation  one could thus make an explicit calculation possible.

The fact that the BD action does not vanish in flat spacetime suggests  that this action  does not only code information about the bulk Einstein Hilbert action but also contains  extra boundary information in general.  One might then be tempted to speculate that it contains {\it all} the relevant boundary information suggesting that the null boundary contribution for an Alexandrov interval in any spacetime always vanishes. Moreover, in the specific case that there is only a single spatial component coming from the joint, that the only boundary contribution is the volume of this $d-2$ sphere.  comes from this  that (\ref{result}) gives the GHY contribution for an Alexandrov interval.

\newpage
